Evolution of solar wind discontinuities in the inner heliosphere

### 2025-10-29 notebooks/02_stats.qmd
docs: methods (#1)

```{julia}
using Dates
using StatsBase, Statistics
using PSPAnisotropy
import Discontinuity as DC
using Discontinuity: var_mapping, compute_params!, filter_low_mva_quality
using Discontinuity: compute_Alfvenicity_params!, Alfven_velocity
using DimensionalData
using DataFrames, DataFramesMeta
using CairoMakie
using Beforerr: _theme_legend, easy_save
using AlgebraOfGraphics
using AlgebraOfGraphics: data
import AlgebraOfGraphics as AoG
using LinearAlgebra
using Accessors
using Unitful
Unitful.preferunits(u"nA", u"km")

# https://github.com/MakieOrg/AlgebraOfGraphics.jl/issues/464
foreach([:log10, :log2, :log]) do f
    @eval import Base: $f
    @eval $f(x::Unitful.Quantity) = $f(ustrip(x))
end
set_Z_theme!()
formats = (:pdf, :png)
```

@@

taus = Second.(2 .^ (1:6))
# taus = Second.(2 .^ (0:6))
df = workload(taus)

@@

@chain df begin
    DC.compute_Alfvenicity_params!()
    DC.compute_params!()
    @transform! @astable begin
        :Δ𝐁 = @. norm(:B_us - :B_ds)
        :Δ𝐁_fit = @. 2 * passmissing(A_lens)(:model) |> abs
        :Δ𝐁_norm = :Δ𝐁 ./ :B_mag
        :Δ𝐁_fit_norm = :Δ𝐁_fit ./ :B_mag
        :B_n_mva_norm = abs.(:B_n_mva ./ :B_mag)
        :duration = _float64.(ustrip.(:duration .|> u"s"))
    end
    DC.classify!(:B_n_mva_norm)
end

@@

subset_𝐧(df, 𝐧) = 𝐧 == :mva ? filter_low_mva_quality(df) : df
```

@@

fs = map([:mva]) do 𝐧
    figure = (; size=(450, 600))
    𝒎 = var_mapping(; 𝐧)

@@

    tdf = @chain df begin
        @subset(.!isnan.(:L_n_mva_norm); view=true)
        subset_𝐧(𝐧)
        dropmissing(:L_n_mva_norm)
    end

    vars = [𝒎.l_log, 𝒎.l_norm_log, 𝒎.j_log, 𝒎.j_norm_log, 𝒎.Δ𝐁, 𝒎.Δ𝐁_norm_log]
    labels = ["Thickness", "Normalized thickness", "Current density", "Normalized current density", "|Δ𝐁|", "|Δ𝐁|/⟨|𝐁|⟩"]
    # append!(vars, [ :Δ𝐁_fit => log10 => "Log Δ𝐁_fit", :Δ𝐁_fit_norm => log10 => "Log Δ𝐁_fit_norm"])
    # append!(labels, ["Δ𝐁_fit", "Δ𝐁_fit_norm"])

    plt = AoG.data(tdf) *
          mapping(vars, layout=AoG.dims(1) => renamer(labels)) *
          mapping(color=:id, linestyle=:enc) *
          AoG.density(; datalimits) * visual(Lines)
    f = draw(plt, scales(Layout=(; palette=wrapped(cols=2))); axis=(; yscale=log10), figure, legend=(; position=:top))
    hide_facet_labels!(f)
    rowgap!(f.figure.layout, 1)
    easy_save("$fname-$𝐧"; formats, force=true)
end
fs[1]
```

@@

```{julia}
let df = dropmissing(df, :duration)
    layer = AoG.density(; datalimits=x -> quantile(x, [0.02, 0.99])) * visual(Lines)
    # layer = AoG.histogram(; datalimits=x -> quantile(x, [0.02, 0.99]), bins = 0:0.3:30) * mapping(col=:id)
    plt = data(df) * mapping(:duration, linestyle=:enc, color=:id) * layer
    plt = plt + mapping(2 .^ (1:4)) * visual(VLines, color=:gray, linestyle=:dash)
    # plt *= (mapping(col=direct("col1")) + mapping(color=:tau, col=direct("col2")))
    # scales(Col=(; legend=false))
    draw(plt, axis=(; yscale=log10))
    easy_save("duration_dist"; formats)
end
```

@@

```{julia}
let df = dropmissing(df, :duration)
    layer = AoG.density(; datalimits=x -> quantile(x, [0.02, 0.99])) * visual(Lines)
    plt = data(df) * mapping(:duration, linestyle=:enc, color=:id) * layer
    plt *= mapping(layout=:tau)
    draw(plt, axis=(; yscale=log10), facet=(; linkxaxes=:none, linkyaxes=:none))
    easy_save("duration_dist_tau"; formats)
end
```

@@

for 𝐧 in [:mva]
    view = true
    𝒎 = var_mapping(; 𝐧)
    tdf = @chain df begin
        dropmissing(:L_n_mva)
        @transform(:L_n_mva = Float64.(ustrip.(:L_n_mva)), :J_m_max_mva = Float64.(ustrip.(:J_m_max_mva)))
        subset_𝐧(𝐧)
    end

@@

    spec1 = plt * mapping(𝒎.l, 𝒎.j)
    spec2 = plt * mapping(𝒎.l_norm, 𝒎.j_norm)
    B_l_spec = plt * mapping(𝒎.Δ𝐁_norm, 𝒎.l_norm) # Not strong correlation
    B_j_spec = plt * mapping(𝒎.Δ𝐁_norm, 𝒎.j_norm) # Not strong correlation

    grid1 = draw!(fig[1, 1], spec1; axis)
    grid2 = draw!(fig[2, 1], spec2; axis)
    grid3 = draw!(fig[3, 1], B_l_spec; axis)
    grid4 = draw!(fig[4, 1], B_j_spec; axis)

    hide_facet_labels!.((grid2, grid3, grid4))

    add_labels!(fig[1:4, 1]; position=Left(), padding=(0, 37, 0, 0))
    display(fig)
    easy_save("joint_properties-$𝐧"; formats, force=true)
end
```

@@

for 𝐧 in [:mva], subset in (true, false)
    normalization = :column
    𝒎 = var_mapping(; 𝐧)
    tdf = dropmissing(df, :B_n_mva_norm)
    tdf = subset ? subset_𝐧(tdf, 𝐧) : tdf
    plt = AoG.data(tdf) * mapping(col=:id => layout_func)
    B_n_ω_spec = plt * mapping(𝒎.bn, 𝒎.ω_in) * fhist(; binedges=(0:0.03:0.99, 0:6:180), normalization)
    f = draw(B_n_ω_spec)
    colgap!(f.figure.layout, 10)

    den_base = AoG.data(tdf) * mapping(color=:id => layout_func) * AoG.density(; datalimits) * visual(Lines)
    ff = f.figure[0, 1:3]
    grid1 = draw!(ff[1, 1], den_base * mapping(𝒎.ω_in))
    grid2 = draw!(ff[1, 2], den_base * mapping(𝒎.bn); axis=(; ylabel="", yscale=log10))
    legend!(ff[1, 1], grid1; tellheight=false, tellwidth=false, halign=:right, valign=:top)
    rowgap!(f.figure.layout, 4)
    add_labels!([ff[1, 1], ff[1, 2], f.figure[1, 1:end]]; position=Left(), padding=(0, 40, 0, -5))
    easy_save("B_n_ω_subset=$subset"; formats, force=true)
end
```

@@

    Q_limit = (-0.992, 0.992) # mainly for plotting so not show (-1, 1) in ticks
    ΔV_spec = plt * mapping(𝒎.ΔV_ratio, 𝒎.ΔV_cosθ) * AoG.density(; datalimits=((0, 1.1), (-1, 1)))
    duration_Q_spec = plt * mapping(:duration, 𝒎.Q) * AoG.density(; datalimits=((0, 3.1), Q_limit))
    σc_Q_spec = plt * mapping(:σ_c => L"σ_c", 𝒎.Q) * AoG.density()
    σr_Q_spec = plt * mapping(:σ_r => L"σ_r", 𝒎.Q) * AoG.density(; datalimits=(Q_limit, Q_limit))

@@

    rowgap!(fig.layout, 4)
    for g in (grid3, grid4)
        ax = g[1].axis
        ax.xlabelpadding[] = 1
        ax.xlabelsize[] = 16
    end
    add_labels!(fig[1:4, 1]; position=Left(), padding=(0, 0, 0, -5))
    hide_facet_labels!.((grid2, grid3, grid4))
    easy_save("Q_sonnerup_joint_dist_den"; formats, force=true)
    display(fig)
end
```

>>>>

```{julia}
using Dates
using StatsBase, Statistics
using PSPAnisotropy
import Discontinuity as DC
using Discontinuity: var_mapping, compute_params!, filter_low_mva_quality
using Discontinuity: compute_Alfvenicity_params!, Alfven_velocity
using DimensionalData
using DataFrames, DataFramesMeta
using CairoMakie
using Beforerr: _theme_legend, easy_save, hide_facet_labels!, add_labels!
using AlgebraOfGraphics
using AlgebraOfGraphics: data
import AlgebraOfGraphics as AoG
using LinearAlgebra
using Accessors
using Unitful
Unitful.preferunits(u"nA", u"km")

includet("../src/plot.jl")

# https://github.com/MakieOrg/AlgebraOfGraphics.jl/issues/464
foreach([:log10, :log2, :log]) do f
    @eval import Base: $f
    @eval $f(x::Unitful.Quantity) = $f(ustrip(x))
end
set_Z_theme!()
```

@@

taus = Second.(2 .^ (1:6))
# taus = Second.(2 .^ (0:6))
wdf = workload(taus)

@@

df = @chain wdf begin
    DC.compute_Alfvenicity_params!()
    DC.compute_params!()
    @transform @astable begin
        :Δ𝐁 = @. norm(:B_us - :B_ds)
        :Δ𝐁_fit = @. 2 * passmissing(A_lens)(:model) |> abs
        :Δ𝐁_norm = :Δ𝐁 ./ :B_mag
        :Δ𝐁_fit_norm = :Δ𝐁_fit ./ :B_mag
        :B_n_mva_norm = abs.(:B_n_mva ./ :B_mag)
        :duration = _float64.(ustrip.(:duration .|> u"s"))
        :L_n_mva = _float64.(ustrip.(:L_n_mva))
        :J_m_max_mva = _float64.(ustrip.(:J_m_max_mva))
        :L_n_cross = _float64.(ustrip.(:L_n_cross))
        :J_m_max_cross = _float64.(ustrip.(:J_m_max_cross))
    end
    DC.classify!(:B_n_mva_norm)
end

@@

subset_𝐧(df, 𝐧) = 𝐧 == :mva ? filter_low_mva_quality(df) : df
const datalimits = x -> quantile(x, [0.01, 0.99])
```

@@

update_theme!(;
    Legend=(; framevisible=false, padding=(0, 0, 0, 0)),
)

function plot_properties_hist(df, 𝐧)
    figure = (; size=(450, 600))
    𝒎 = var_mapping(; 𝐧)

@@

    l_sym = first(𝒎.l_norm)
    tdf = @chain df begin
        @subset(.!isnan.($l_sym); view=true)
        subset_𝐧(𝐧)
        dropmissing(l_sym)
    end

    l_vars = [𝒎.l_log, 𝒎.l_norm_log]
    j_vars = [𝒎.j_log, 𝒎.j_norm_log]
    B_vars = [𝒎.Δ𝐁, 𝒎.Δ𝐁_norm_log]
    l_labels = ["Thickness", "Normalized thickness"]
    j_labels = ["Current density", "Normalized current density"]
    B_labels = ["|Δ𝐁|", "|Δ𝐁|/⟨|𝐁|⟩"]

    facet = (; linkxaxes=:none, linkyaxes=:none)
    layer = mapping(color=:id, linestyle=:enc) *
            AoG.density(; datalimits) * visual(Lines)
    axis = (; yscale=log10)

    plt = AoG.data(tdf) * mapping(l_vars, layout=AoG.dims(1)) * layer
    f = draw(plt; axis, figure, legend=(; position=:top))
    ylims!.(getproperty.(f.grid, :axis), 10^(-1.6), 10^(0.1))

    gl2 = GridLayout(f.figure[2, 1:end])
    f2 = draw!(gl2, AoG.data(tdf) * mapping(j_vars, layout=AoG.dims(1)) * layer; axis)

    gl3 = GridLayout(f.figure[3, 1:end])
    # no need to subset and dropmissing
    f3 = draw!(gl3, AoG.data(df) * mapping(B_vars, layout=AoG.dims(1)) * layer; axis)

    hide_facet_labels!.((f, f2, f3))
    rowgap!(f.figure.layout, 1)
    add_labels!(f.figure; position=Left(), padding=(20, -20, 0, 0))
    colgap!.((f.figure.layout, gl2, gl3), 1)

    return f
end

fs = map([:mva, :cross]) do 𝐧
    plot_properties_hist(df, 𝐧)
    # subset df by plasma beta
    # plot_properties_hist(@rsubset(df, :β < 1), 𝐧)
    # plot_properties_hist(@rsubset(df, :β > 3), 𝐧)
    # easy_save("$fname-$𝐧"; force=true)
    easy_save("$fname-$𝐧")
end
fs[1]
```

![properties_hist-mva](../figures/properties_hist-mva.png)
![properties_hist-cross](../figures/properties_hist-cross.png)

@@

```{julia}
let tdf = dropmissing(df, :duration)
    psp_df = @chain tdf begin
        @rsubset(:id == "PSP"; view=true)
        @transform(:duration = :duration .* 4, :id = "PSP (4x)")
    end
    fdf = vcat(tdf, psp_df)
    axis = (; yscale=log10)
    layer = mapping(:duration) * AoG.density(; datalimits=x -> quantile(x, [0.01, 0.99])) * visual(Lines)
    plt = data(fdf) * mapping(color=:id) * layer
    plt = plt + mapping(2 .^ (0:5)) * visual(VLines, color=:gray, linestyle=:dash)
    fg = draw(plt; axis)

    easy_save("duration_dist")
end
```

![](../figures/duration_dist.png)

@@

```{julia}
let df = dropmissing(df, :duration)
    layer = AoG.density(; datalimits=x -> quantile(x, [0.02, 0.99])) * visual(Lines)
    plt = data(df) * mapping(:duration, linestyle=:enc, color=:id) * layer
    plt *= mapping(layout=:tau)
    draw(plt, axis=(; yscale=log10), facet=(; linkxaxes=:none, linkyaxes=:none))
    easy_save("duration_dist_tau")
end
```

@@

for 𝐧 in [:mva, :cross]
    view = true
    𝒎 = var_mapping(; 𝐧)

    l_sym = first(𝒎.l_norm)

    tdf = @chain df begin
        dropmissing(l_sym)
        subset_𝐧(𝐧)
    end

@@

    spec1 = plt * mapping(𝒎.l, 𝒎.j)
    spec2 = plt * mapping(𝒎.l_norm, 𝒎.j_norm)
    B_l_spec = plt * mapping(𝒎.l_norm, 𝒎.Δ𝐁_norm) # Not strong correlation
    B_j_spec = plt * mapping(𝒎.j_norm, 𝒎.Δ𝐁_norm) # Not strong correlation

    d_lims = (10^(-2.5), 10^3.5)
    j_lims = (10^(-3.6), 10^2.4)
    B_lims = (10^(-3), 10^3)

    grid1 = draw!(fig[1, 1], spec1; axis)
    grid2 = draw!(fig[2, 1], spec2; axis=(; axis..., limits=(d_lims, j_lims)))
    grid3 = draw!(fig[3, 1], B_l_spec; axis=(; axis..., limits=(d_lims, nothing)))
    grid4 = draw!(fig[4, 1], B_j_spec; axis=(; axis..., limits=(j_lims, nothing)))

    hide_facet_labels!.((grid2, grid3, grid4))
    rowgap!(fig.layout, 4)

    add_labels!([fig[1, 1], fig[2, 1], fig[3, 1], fig[4, 1]]; position=Left(), padding=(0, 37, 0, 0))
    display(fig)
    # easy_save("joint_properties-$𝐧"; force=true)
    easy_save("joint_properties-$𝐧")
end
```

![](../figures/joint_properties-mva.png)
![](../figures/joint_properties-cross.png)

@@

for 𝐧 in [:mva, :cross], subset in (true)
    normalization = :column
    𝒎 = var_mapping(; 𝐧)
    B_sym = Symbol(:B_n_, 𝐧)
    tdf = dropmissing(df, B_sym)
    # tdf = subset ? subset_𝐧(tdf, 𝐧) : tdf
    tdf = subset_𝐧(tdf, 𝐧)
    plt = AoG.data(tdf) * mapping(col=:id => layout_func)
    B_n_ω_spec = plt * mapping(𝒎.bn, 𝒎.ω_in) * fhist(; binedges=(0:0.03:0.99, 0:6:180), normalization)
    f = draw(B_n_ω_spec)
    colgap!(f.figure.layout, 10)

    den_base = AoG.data(tdf) * mapping(color=:id => layout_func) * AoG.density(; datalimits) * visual(Lines)
    ff = f.figure[0, 1:3]
    grid1 = draw!(ff[1, 1], den_base * mapping(𝒎.ω_in))
    grid2 = draw!(ff[1, 2], den_base * mapping(𝒎.bn); axis=(; ylabel="", yscale=log10))
    legend!(ff[1, 1], grid1; tellheight=false, tellwidth=false, halign=:right, valign=:top)
    rowgap!(f.figure.layout, 4)
    add_labels!([ff[1, 1], ff[1, 2], f.figure[1, 1:end]]; position=Left(), padding=(0, 40, 0, -5))
    # easy_save("B_n_ω_subset=$subset"; force=true)
    easy_save("B_n_ω_-$𝐧-subset=$subset")
end
```

for 𝐧 in [:mva], subset in (true, false)
    normalization = :column
    𝒎 = var_mapping(; 𝐧)
    tdf = dropmissing(df, :B_n_mva_norm)
    tdf = subset ? subset_𝐧(tdf, 𝐧) : tdf
    plt = AoG.data(tdf) * mapping(col=:id => layout_func)
    duration_spec = plt * mapping(:duration, 𝒎.ω_in)
    f = draw(duration_spec)
    display(f)
end
```

@@

    Q_limit = (-0.992, 0.992) # mainly for plotting so not show (-1, 1) in ticks
    ΔV_spec = plt * mapping(𝒎.ΔV_ratio, 𝒎.ΔV_cosθ) * AoG.density(; datalimits=((0, 1.1), (-1, 1)))
    # duration_Q_spec = plt * mapping(:duration, 𝒎.Q) * AoG.density(; datalimits=((0, 3.1), Q_limit))
    duration_Q_spec = plt * mapping(:duration => log10 => "Log Duration (s)", 𝒎.Q) * AoG.density(; datalimits=((-0.9, 1.1), Q_limit))
    # duration_Q_spec = plt * mapping(:duration => log10, 𝒎.Q) * fhist(; binedges=(-0.9:0.1:1.1, -0.99:0.1:0.99), normalization=:column)
    σc_Q_spec = plt * mapping(:σ_c => L"σ_c", 𝒎.Q) * AoG.density()
    σr_Q_spec = plt * mapping(:σ_r => L"σ_r", 𝒎.Q) * AoG.density(; datalimits=(Q_limit, Q_limit))

@@

    rowgap!(fig.layout, 4)
    for g in (grid3, grid4)
        ax = g[1].axis
        ax.xlabelpadding[] = 1
        ax.xlabelsize[] = 16
    end
    add_labels!(fig[1:4, 1]; position=Left(), padding=(0, 0, 0, -5))
    hide_facet_labels!.((grid2, grid3, grid4))
    easy_save("Q_sonnerup_joint_dist_den")
end
```

### 2025-10-29 notes.qmd
docs: methods (#1)

##

![](images/paste-20.png)

@@

>>>>

@@

<!-- ::: {#fig-plasma-adiabatic-evolution .column-page}
![](figures/evolution/plasma-adiabatic-evolution_e2-DSCOVR-next.svg)

![](figures/evolution/plasma-adiabatic-evolution_e2-DSCOVR.svg)

Comparison of plasma dimensionless quantities (plasma beta and normalized plasma speed by Alfven speed) between PSP (red) and DSCOVR (blue). Yellow dots are predicted plasma beta and normalized speed using formulary from @perroneRadialEvolutionSolar2019.
::: -->

<!-- sensing solar wind emitted from the same region on the solar surface, with the caveat that dynamical processes at the chromosphere–corona interface have not dramatically altered the solar footpoint within the time corresponding to the radial propagation from the measurement of one spacecraft to that of the other. -->

<!-- OUTDATED
To achieve this, we utilize a configuration where two spacecraft are radially aligned in the Heliographic (HG) frame, with the longitude fixed on the Sun. It is important to note that the plasma parcels observed by the two spacecraft may differ in composition, as solar wind particles propagate at varying velocities. For instance, protons often exhibit a beam in their velocity distribution function, while the thermal speed of electrons is much higher than their bulk speed. Nonetheless, observing plasma that passes through both spacecraft offers the best opportunity to study the evolution of solar wind discontinuities with minimal influence from other processes during propagation. It is also expected that these plasma parcels evolve due to heating and other processes.

The plasma parcels in question are large-scale structures. As the scale size decreases below the mesoscale [@viallMesoscaleStructureSolar2021] (with radial sizes ranging from approximately 5 × 10³ km to 10⁷ km), it becomes increasingly unlikely to identify the same structure, depending on the latitude difference between the two spacecraft. Furthermore, although large-scale structures (especially density enhancements) can be identified as passing through both spacecraft [@berriotIdentificationSinglePlasma2024], it is nearly impossible to detect kinetic-scale structures, such as solar wind discontinuities (with time scales on the order of seconds), by both spacecraft simultaneously. Therefore, our study focuses on the statistical properties of solar wind discontinuities within the "same" large-scale structures that pass through both spacecraft. 
-->
